Design and Implementation of Probabilistic Programming Language Anglican

### 2016-07-07 paper/paper.tex
typos fixed

\begin{abstract}
    Anglican is a probabilistic programming system designed to interoperate
    with Clojure and other JVM languages. We introduce the
	programming language Anglican, outline our design choices,
	and discuss in depth the implementation of the Anglican language
	and runtime, including macro-based compilation, extended
	CPS-based evaluation model, and functional representations for
	probabilistic paradigms, such as distribution,
	random processes, and inference algorithms.

	We show that a probabilistic functional language can be
	implemented efficiently and integrated tightly with a
	conventional functional language with only moderate
	computational overhead. We also demonstrate how advanced
	probabilistic modeling concepts are mapped naturally
	to a functional representation.
\end{abstract}

@@

Internally, an Anglican query is represented by a computation in
\textit{continuation passing style} (CPS), and inference algorithms
exploit the CPS structure of the code to intercept probabilistic
operations in an algorithm-specific way\footnote{\cite{GS15}
also describe a CPS-based implementation of a probabilistic
programming language.}. Among the available
inference algorithms there are Particle Cascade~\cite{PWD+14},
Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
Inference on Anglican queries generates a lazy sequence of
samples, which can be processed asynchronously in Clojure
code for analysis, integration, and decision making.

@@

A probabilistic program, or query, mostly runs deterministic
code, except for certain checkpoints, in which probabilities are
involved, and normal, linear execution of the program is
disrupted. In Anglican and similar languages there are two
types of such checkpoints:
\begin{itemize}
    \item drawing a value from a random source (\texttt{sample});
    \item conditioning the posterior distribution by
        conditioning a computed value on a random source
        (\texttt{observe}).
\end{itemize}

@@

Compilation of Anglican into Clojure relies on the Clojure
\textit{macro} facility. However, the compilation algorithm is
implemented as a library of \textit{functions} in namespace
\texttt{anglican.trap}, which are invoked by macros. Namespace
\texttt{anglican.trap-test} contains many transformation tests,
which both help assure proper functioning of the transformation
functions, and serve as illustrative examples of individual
transformations, such as of functions, flow control, and
probabilistic contstructs.

The CPS transformation is organized in top-down manner. The
top-level function is  \texttt{cps-of-expression}, which
receives an expression and a continuation, and returns the
expression in the CPS form, with the computed result passed to
the continuation. A continuation accepts \texttt{two arguments}:
\begin{itemize}
    \item the computed value;
    \item the internal state, bound to the local variable
        \texttt{\$state} in every lexical scope.
\end{itemize}
The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithm. \texttt{\$state}
is a Clojure hash map, and the map entries are
algorithm-dependent. Except for transformation of
\texttt{mem}, the memoization form, CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

@@

Similar to distributions, \textit{random processes}, related to
so called `exchangeable random procedures'\footnote{However, random sequences
generated by Anglican random processes are not required to be
exchangeable.}, are defined using
macro \texttt{defproc} and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. The protocol
has twho methods --- \texttt{produce}, which returns a
distribution object, and \texttt{absorb}, which returns a new
random process object obtained by absorbing a value drawn from
the process. While distributions are random and non-functional,
random processes are deterministic and functional, hence
\texttt{produce} and \texttt{absorb} are called directly and do
not have corresponding special forms in Anglican.

>>>>

\begin{abstract}
    Anglican is a probabilistic programming system designed to interoperate
    with Clojure and other JVM languages. We introduce the
	programming language Anglican, outline our design choices,
	and discuss in depth the implementation of the Anglican language
	and runtime, including macro-based compilation, extended
	CPS-based evaluation model, and functional representations for
	probabilistic paradigms, such as a distribution,
	a random process, and an inference algorithm.

	We show that a probabilistic functional language can be
	implemented efficiently and integrated tightly with a
	conventional functional language with only moderate
	computational overhead. We also demonstrate how advanced
	probabilistic modeling concepts are mapped naturally
	to the functional foundation.
\end{abstract}

@@

Internally, an Anglican query is represented by a computation in
\textit{continuation passing style} (CPS)~\cite{AJ89}, and inference algorithms
exploit the CPS structure of the code to intercept probabilistic
operations in an algorithm-specific way\footnote{\cite{GS15}
also describe a CPS-based implementation of a probabilistic
programming language.}. Among the available
inference algorithms there are Particle Cascade~\cite{PWD+14},
Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
Inference on Anglican queries generates a lazy sequence of
samples, which can be processed asynchronously in Clojure
code for analysis, integration, and decision making.

@@

A probabilistic program, or query, mostly runs deterministic
code, except for certain checkpoints, in which probabilities are
involved, and normal, linear execution of the program is
disrupted. In Anglican and similar languages there are two types of such checkpoints:
\begin{itemize} \item drawing a value from a random source (\texttt{sample});
    \item conditioning a computed value on a random source
        (\texttt{observe}).
\end{itemize}

@@

Compilation of Anglican into Clojure relies on the Clojure
\textit{macro} facility. However, the compilation algorithm is
implemented as a library of \textit{functions} in namespace
\texttt{anglican.trap}, which are invoked by macros. Namespace
\texttt{anglican.trap-test} contains many transformation tests,
which both help assure proper functioning of the transformation
functions, and serve as illustrative examples of individual
transformations, such as of functions, flow control, and
probabilistic constructs.

The CPS transformation is organized in top-down manner. The
top-level function is  \texttt{cps-of-expression}, which
receives an expression and a continuation, and returns the
expression in the CPS form, with the computed result passed to
the continuation. A continuation accepts \textit{two} arguments:
\begin{itemize}
    \item the computed value;
    \item the internal state, bound to the local variable
        \texttt{\$state} in every lexical scope.
\end{itemize}
The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithm. \texttt{\$state}
is a Clojure hash map, and the map entries are
algorithm-dependent. Except for transformation of
\texttt{mem}, the memoization form, CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

@@

Similar to distributions, \textit{random processes}, related to
so called `exchangeable random procedures'\footnote{However, random sequences
generated by Anglican random processes are not required to be
exchangeable.}, are defined using
macro \texttt{defproc} and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. The protocol
has two methods --- \texttt{produce}, which returns a
distribution object, and \texttt{absorb}, which returns a new
random process object obtained by absorbing a value drawn from
the process. While distributions are random and non-functional,
random processes are deterministic and functional, hence
\texttt{produce} and \texttt{absorb} are called directly and do
not have corresponding special forms in Anglican.
